Empirical pipeline

### 2024-06-03 Analysis/01_Simulating_spaces_and_stressors.Rmd
Cleaned up analysis folder

We can use the package `dads` that's in development for simulating the trait spaces (note that the package is still pretty unstable but the simple tasks here should not change).
Later, we can use the `dispRity::reduce.space` package to apply stressors to the data.

@@

## Paper figure

```{r}
plot.text.box <- function(my_text, ...) {
  plot(1,1, xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = "white", pch = 21)
  text(c(1,1), my_text, ...)
}

>>>>

We can use the package `treats` that's in development for simulating the trait spaces (note that the package is still pretty unstable but the simple tasks here should not change).
Later, we can use the `dispRity::reduce.space` package to apply stressors to the data.

@@

## Paper figure 1

```{r, eval = FALSE}
plot.text.box <- function(my_text, ...) {
  plot(1,1, xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = "white", pch = 21)
  text(c(1,1), my_text, ...)
}

### 2024-06-03 Analysis/02_Simulation_pipeline.Rmd
Cleaned up analysis folder

```{r}
## Loading the functions
source("../Functions/melodic.rao.R")
source("../Functions/make.simulation.pipeline.R")
source("../Functions/simulation.pipeline.R")
source("../Functions/dbFD.R")
```

@@

There are options to

```{r, eval = TRUE}
## Running the full optimisation
results <- optim.replicate(input.fun = simulate.facilitation,
                           diagnose = var,
                           summarise = optim.summarise,
                           minimum = 20, maximum = 170,
                           stop.variance = 0.01,
                           verbose = TRUE,
                           bkp.path = "../Data/Processed/",
                           bkp.name = "sim_facilitation.rda")
```
> Note that you can use the parallel version by adding the argument `parallel = TRUE` that is automatised using the package `future` and `future.apply` which makes the whole process run on all your available cores but is way less verbose. Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.

>>>>

```{r}
## Loading the functions
source("../Functions/melodic.rao.R")
source("../Functions/make.simulation.pipeline.R")
source("../Functions/simulation.pipeline.R")
source("../Functions/analyse.replicates.R")
source("../Functions/dbFD.R")
```

@@

```{r, eval = TRUE}
## Running the full optimisation
results <- optim.replicate(input.fun = simulate.facilitation,
                           diagnose = var,
                           summarise = optim.summarise,
                           minimum = 20, maximum = 170,
                           stop.variance = 0.01,
                           verbose = TRUE,
                           bkp.path = "../Data/Processed/",
                           bkp.name = "sim_facilitation.rda")
```

> Note that you can use the parallel version by adding the argument `parallel = TRUE` that is automatised using the package `future` and `future.apply` which makes the whole process run on all your available cores but is way less verbose. Without using the parallel version, it takes roughly 10 hours (overnight) to run the analyses on an old 2.2Ghz 8GB core.

You can run this pipeline on a cluster using/modifying the `cluster_script_example.Rscript` file in `Analysis/`.

### 2024-06-03 Analysis/03_Empirical_pipeline.Rmd
Cleaned up analysis folder

```{r}
## Loading the functions
source("../Functions/melodic.rao.R")
source("../Functions/make.simulation.pipeline.R")
source("../Functions/simulation.pipeline.R")
```

@@

You can run the pipeline on the empirical data if you have the following elements ready:

 * A trait space (a `"matrix"`) with the rows as species and the columns as dimensions (make sure it has rownames for avoiding any problems down the line).
 * A list of species to remove for each sub-sampling the list should be list of named `logical` vectors (where for example species that should be excluded - e.g. fossils are marked as `FALSE` and the ones to include as `TRUE`).

Here are two dummy examples but you can replace them with your empirical data:

```{r}
## Making a dummy trait space for 200 species (3D)
my_traitspace <- dispRity::space.maker(elements = 200, dimensions = 3,
                                       distribution = rnorm,
                                       elements.names = 1:200)
## This is what it looks like
head(my_traitspace)

## Making a dummy set of sub-sampling
my_subsamples <- list(sample(c(TRUE, FALSE), 200, replace = TRUE),
                      sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.75, 0.25)),
                      sample(c(TRUE, FALSE), 200, replace = TRUE, prob = c(0.90, 0.10)))
## Adding the elements names
my_subsamples <- lapply(my_subsamples, function(x) {names(x) <- 1:200 ; return(x)})

## This is what it looks like
str(my_subsamples)
```

@@

```{r}
## Creating the function for running the empirical pipeline
empirical.pipeline <- make.simulation.pipeline(sim.data = my_traitspace,
                                               remove   = my_subsamples)
```

@@

```{r, eval = TRUE}
## Running the full optimisation
results <- optim.replicate(input.fun = empirical.pipeline,
                           diagnose = var,
                           summarise = optim.summarise,
                           minimum = 20, maximum = 170,
                           stop.variance = 0.0001,
                           verbose = TRUE,
                           bkp.path = "../Data/Processed/",
                           bkp.name = "empirical_data.rda")
```

@@

## Checking the results

We can then visualise the "convergence" of the results using the `analyse.replicate` function:

```{r, eval = FALSE}
source("../Functions/analyse.replicates.R")
## Check the change in variance for all 120 parameters
analyse.replicates("../Data/Processed/sim_filtering.rda") #140 -> #170
analyse.replicates("../Data/Processed/sim_equalizing.rda") #155
analyse.replicates("../Data/Processed/sim_competition.rda") #126
analyse.replicates("../Data/Processed/sim_facilitation.rda") #163
analyse.replicates("../Data/Processed/empirical_data.rda") #163
```

>>>>

```{r}
## Loading the functions
source("../Functions/melodic.rao.R")
source("../Functions/make.simulation.pipeline.R")
source("../Functions/simulation.pipeline.R")
source("../Functions/analyse.replicates.R")
source("../Functions/dbFD.R")
```

@@

To run the simulations on an empirical data set you'll need to:

 1. load an empirical traitspace
 2. load a list of species to remove
 3. generate the pipeline as for the simulated data

Here's the pipeline applied to the Hawaiian birds dataset/

```{r}
## global trait space 
empirical_traitspace <- read.csv('../Data/Raw/trait-space-5d-prehistoric-community.csv', row.names = 1)

## Species lists
historic_species <- rownames(read.csv('../Data/Raw/historic-species-list-minus-a-flammeus.csv', row.names = 1))
extant_species <- rownames(read.csv('../Data/Raw/present-species-list-minus-a-flammeus.csv', row.names = 1))

## Traitspace groups
extant_group <- rownames(empirical_traitspace) %in% extant_species ## 37 species present
historic_group <- rownames(empirical_traitspace) %in% historic_species ## 63 species present
groups <- list("extant" = extant_group, "historic" = historic_group)
```

@@

```{r}
## Creating the function for running the empirical pipeline
empirical.pipeline <- make.simulation.pipeline(sim.data = empirical_traitspace,
                                               remove = groups)
```

@@

```{r, eval = TRUE}
## Running the full optimisation
results <- optim.replicate(input.fun = empirical.pipeline,
                           diagnose = var,
                           summarise = optim.summarise,
                           minimum = 20, maximum = 170,
                           stop.variance = 0.01,
                           verbose = TRUE,
                           bkp.path = "../Data/Processed/",
                           bkp.name = "empirical_data.rda")
```

@@
